The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2021-01-19 journal paper/content/sim_details.tex
Space-saving.

\xhdr{MAB algorithms.} In abstract terms, we posit three types of technology, from ``low" to ``medium" to ``high". Concretely, we consider three essential classes of bandit algorithms: ones that never explicitly explore (\emph{greedy algorithms}), ones that explore without looking at the data (\emph{exploration-separating algorithms}), and ones where exploration gradually zooms in on the best arm (\emph{adaptive-exploration algorithms}). In the absence of competition, these classes are fairly well-understood: greedy algorithms are terrible for a wide variety of problem instances, exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances, and adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances (see Appendix~\ref{app:bg} for more details).

>>>>

\xhdr{MAB algorithms.} In abstract terms, we posit three types of technology, from ``low" to ``medium" to ``high". Concretely, we consider three essential classes of bandit algorithms: ones that never explicitly explore (\emph{greedy algorithms}), ones that explore without looking at the data (\emph{exploration-separating algorithms}), and ones where exploration gradually zooms in on the best arm (\emph{adaptive-exploration algorithms}). In the absence of competition, these classes are fairly well-understood: greedy algorithms are terrible for a wide variety of problem instances, exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances, and adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances (see Appendix~\ref{app:bg}).

### 2021-01-19 journal paper/ec19paper/content/barriers.tex
update figures
Space-saving.

\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
\hline
  & \multicolumn{3}{c||}{Reputation advantage (only)}
  & \multicolumn{3}{c|}{Data advantage (only)}\\
\hline
& $\TS$  & $\DEG$  & $\DG$
& $\TS$  & $\DEG$  & $\DG$
\\\hline
$\TS$
    & \makecell{\textbf{0.021}$\pm$0.009}
    & \makecell{\textbf{0.16}$\pm$0.02}
    & \makecell{\textbf{0.21} $\pm$0.02}
    %%
    & \makecell{\textbf{0.0096}$\pm$0.006}
    & \makecell{\textbf{0.11}$\pm$0.02}
    & \makecell{\textbf{0.18}$\pm$0.02}
    \\ \hline
$\DEG$
    & \makecell{\textbf{0.26}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02}
    & \makecell{\textbf{0.26}$\pm$0.02}
    %%
    & \makecell{\textbf{0.073}$\pm$0.01}
    & \makecell{\textbf{0.29}$\pm$0.02}
    & \makecell{\textbf{0.25}$\pm$0.02}
    \\ \hline
$\DG$
    & \makecell{\textbf{0.34}$\pm$0.03}
    & \makecell{\textbf{0.4}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    %%
    & \makecell{\textbf{0.15}$\pm$0.02}
    & \makecell{\textbf{0.39}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    \\\hline
\end{tabular}
\caption{Data advantage vs. reputation advantage experiment, on Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the {\bf row}) and the incumbent's algorithm (the {\bf column}). The cell specifies the entrant's market share for the rounds in which hit was present: the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{barrier_exp}
\end{table*}

@@

\begin{finding}\label{barrier-find}
\textit{\\
(a) Data advantage and reputation advantage alone are large, across all algorithms and MAB instances. \\(b) The data advantage is larger than the reputation advantage when the incumbent chooses \Thompson. \\(c) The two advantages are similar in magnitude when the incumbent chooses \DynamicEpsGreedy or \DynamicGreedy.
}
\end{finding}

>>>>

\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
\hline
  & \multicolumn{3}{c||}{Reputation advantage (only)}
  & \multicolumn{3}{c|}{Data advantage (only)}\\
\hline
& $\TS$  & $\DEG$  & $\DG$
& $\TS$  & $\DEG$  & $\DG$
\\\hline
$\TS$
    & \makecell{\textbf{0.021}$\pm$0.009}
    & \makecell{\textbf{0.16}$\pm$0.02}
    & \makecell{\textbf{0.21} $\pm$0.02}
    %%
    & \makecell{\textbf{0.0096}$\pm$0.006}
    & \makecell{\textbf{0.11}$\pm$0.02}
    & \makecell{\textbf{0.18}$\pm$0.02}
    \\ \hline
$\DEG$
    & \makecell{\textbf{0.26}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02}
    & \makecell{\textbf{0.26}$\pm$0.02}
    %%
    & \makecell{\textbf{0.073}$\pm$0.01}
    & \makecell{\textbf{0.29}$\pm$0.02}
    & \makecell{\textbf{0.25}$\pm$0.02}
    \\ \hline
$\DG$
    & \makecell{\textbf{0.34}$\pm$0.03}
    & \makecell{\textbf{0.4}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    %%
    & \makecell{\textbf{0.15}$\pm$0.02}
    & \makecell{\textbf{0.39}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    \\\hline
\end{tabular}
\caption{\footnotesize Data advantage vs. reputation advantage experiment, on Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the {\bf row}) and the incumbent's algorithm (the {\bf column}). The cell specifies the entrant's market share for the rounds in which hit was present: the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{barrier_exp}
\end{table*}

@@

\begin{finding}\label{barrier-find}
\textit{
(a) Data advantage and reputation advantage alone are large, across all algorithms and MAB instances. (b) The data advantage is larger than the reputation advantage when the incumbent chooses \Thompson. (c) The two advantages are similar in magnitude when the incumbent chooses \DynamicEpsGreedy or \DynamicGreedy.
}
\end{finding}

### 2021-01-19 journal paper/ec19paper/content/inverted_u.tex
Space-saving.

Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from  monopoly to ``incumbent" to simultaneous entry duopoly to ``late entrant", we find a stylized inverted-U relationship as in Section~\ref{sec:theory-welfare}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DynamicGreedy, \DynamicEpsGreedy and \Thompson. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs, see \cite{DS-arxiv}). \DynamicGreedy is trivially the dominant strategy under monopoly.

@@

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DynamicGreedy wins out for the trivial reason of tie-breaking. The highest levels are simultaneous entry and ``late entrant". We see that \DynamicGreedy is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DynamicGreedy get stronger when the model transitions from simultaneous entry to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DynamicGreedy goes from a weakly dominant strategy to a strictly dominant one. For the Uniform instance, \DynamicGreedy goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the first-mover regime creates strong incentives for \Thompson. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U-expts}.%
\footnote{We consider the monopoly scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}

@@

\begin{finding}
\textit{
Under simultaneous entry, \Eeog values tend to be much smaller than the time horizon $T$.
}
\end{finding}

We also see that the \Eeog values tend to increase as the warm start $T_0$ increases. We conjecture this is because larger $T_0$ tends to be more beneficial for a better algorithm (as it tends to follow a better learning curve). Indeed, we know that the ``effective end of game" in this scenario typically occurs when a better algorithm loses, and helping it delays the loss.

@@

\xhdr{Welfare implications.}
We study the effects of competition on consumer welfare: the total reward collected by the users over time. Rather than welfare directly, we find it more lucid to consider
\emph{market regret}:
\[ \textstyle T\, \max_a \mu(a) - \sum_{t\in [T]} \mu(a_t), \]
where $a_t$ is the arm chosen by agent $t$. This is a standard performance measure in the literature on multi-armed bandits. Note that smaller regret means higher welfare.

We assume that both firms play their respective equilibrium strategies.
% for the corresponding competition level.
As discussed previously, it is
\DynamicGreedy in the monopoly scenario, and
\DynamicGreedy for both firms in simultaneous entry scenario (Finding \ref{find:duopoly}).
For the first-mover scenario, it is \Thompson for the incumbent (Finding \ref{find:temp-monopoly}) and \DynamicGreedy for the entrant (Finding \ref{find:temp-monopoly-entrant}).

>>>>

Our main experiments concern the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from  monopoly to ``incumbent" to ``simultaneous entry" to ``late entrant", we find a stylized inverted-U relationship as in Section~\ref{sec:theory-welfare}. We look for equilibria in the duopoly game, where each firm's choices are limited to \DynamicGreedy, \DynamicEpsGreedy and \Thompson. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. 
%For cleaner results, 
We break ties towards less advanced algorithms, as they tend to have lower adoption costs \citep{DS-arxiv}. \DynamicGreedy is then the dominant strategy under monopoly.

@@

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DynamicGreedy wins out for the trivial reason of tie-breaking. The highest levels are simultaneous entry and ``late entrant". We see that \DynamicGreedy is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DynamicGreedy get stronger when the model transitions from simultaneous entry to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DynamicGreedy goes from a weakly dominant strategy to a strictly dominant. For the Uniform instance, \DynamicGreedy goes from a Nash equilibrium strategy to a weakly dominant.}
Finally, the middle level of competition, ``incumbent" in the first-mover regime creates strong incentives for \Thompson. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U-expts}.%
\footnote{We consider the monopoly scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}

@@

\begin{finding}
\textit{
Under simultaneous entry, \Eeog values tend to be much smaller than $T$.
}
\end{finding}

We also see that the \Eeog values tend to increase as the warm start $T_0$ increases. We conjecture this is because larger $T_0$ tends to be more beneficial for a better algorithm (as it tends to follow a better learning curve). Indeed, we know that the \Eeog in this scenario typically occurs when a better algorithm loses, and helping it merely delays the loss.

@@

\xhdr{Welfare implications.}
We study the effects of competition on consumer welfare: the total reward collected by the users over time. Rather than welfare directly, we find it more lucid to consider
\emph{market regret}:
$ \textstyle T\, \max_a \mu(a) - \sum_{t\in [T]} \mu(a_t), $
where $a_t$ is the arm chosen by agent $t$. This is a standard performance measure in the literature on multi-armed bandits. Note that smaller regret means higher welfare.

We assume that both firms play their respective equilibrium strategies.
% for the corresponding competition level.
As discussed previously, it is
\DynamicGreedy in the monopoly scenario, and
\DynamicGreedy for both firms for simultaneous entry (Finding \ref{find:duopoly}).
For the first-mover scenario, it is \Thompson for the incumbent (Finding \ref{find:temp-monopoly}) and \DynamicGreedy for the entrant (Finding \ref{find:temp-monopoly-entrant}).

### 2021-01-19 journal paper/ec19paper/content/non_greedy_choice.tex
Lots of edits to address Yishay's comments (all except limited rationality).
Space-saving.

\subsection{Non-deterministic choice Model (\HardMaxRandom)}\label{sec:non_greedy}

@@

\begin{finding}\label{find:non_greedy_choice}
\textit{\TS is weakly dominant under \HardMaxRandom, if and only if $\eps T$ is sufficiently large. Moreover, \HardMaxRandom leads to lower variance in market share, compared to \HardMax.}
\end{finding}

>>>>

\subsection{Non-deterministic choice model (\HardMaxRandom)}\label{sec:non_greedy}

@@

\begin{finding}\label{find:non_greedy_choice}
\textit{\Thompson is weakly dominant under \HardMaxRandom, if and only if $\eps T$ is sufficiently large. Moreover, \HardMaxRandom leads to lower variance in market share, compared to \HardMax.}
\end{finding}

### 2021-01-19 journal paper/ec19paper/content/perf_in_iso.tex
pass through alex's edits
space saver
Space-saving.

We start with a pilot experiment in which we investigate each algorithm's performance ``in isolation": in a stand-alone MAB problem without competition. We focus on reputation scores generated by each algorithm. We confirm that algorithms' performance is ordered as we'd expect:
    $\Thompson > \DynamicEpsGreedy > \DynamicGreedy$
for a sufficiently long time horizon. For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. We also plot the trajectory for instantaneous rewards (\emph{not} averaged over the previous time-periods), which provides a better view into algorithm's performance at a given time.%
\footnote{For ``instantaneous reward" at a given time $t$, we report the average (over all \MRVs) of the mean rewards at this time, instead of the average of the \emph{realized} rewards, so as to decrease the noise.}
Figure \ref{prelim_means} shows these trajectories for the Needle-in-Haystack instance; for other MAB instances the plots are similar.

We summarize this finding as follows:

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=0.35]{ec19paper/figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{ec19paper/figures/relative_nih_ts_dg}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\Thompson \geq \DynamicGreedy$. The shaded area display 95\% confidence intervals.}}
\caption{\footnotesize Relative reputation trajectory for $\Thompson$ vs $\DynamicGreedy$, on Uniform instance (left) and Needle-in-Haystack instance (right). Shaded area display 95\% confidence intervals.}
\label{relative_rep_plots}
\end{figure}

This angle allows a more nuanced analysis of reputation costs vs. benefits under competition. Figure \ref{relative_rep_plots} (left) shows the relative reputation trajectory for $\Thompson$ vs $\DynamicGreedy$ for the Uniform instance. The relative reputation is less than $\tfrac12$ in the early rounds, meaning that $\DynamicGreedy$ has a higher reputation score in a majority of the simulations, and more than $\tfrac12$ later on. The reason is the exploration in \Thompson leads to worse decisions initially, but allows for better decisions later. The time period when relative reputation vs. \DynamicGreedy dips below $\tfrac12$ can be seen as an explanation for the competitive disadvantage of exploration. Such period also exists for the Heavy-Tail MAB instance. However, it does not exist for the Needle-in-Haystack instance, see Figure \ref{relative_rep_plots}.%
\footnote{We see two explanations for this: $\Thompson$ identifies the best arm faster for the Needle-in-Haystack instance, and there are no ``very bad" arms which make exploration very expensive in the short term.}

@@

\noindent Uniform and Heavy-tail instance are exploration-disadvantaged, but Needle-in-Haystack is not.

>>>>

We start with a pilot experiment in which we investigate each algorithm's performance ``in isolation": in a stand-alone MAB problem without competition. We focus on reputation scores generated by each algorithm. We confirm that algorithms' performance is ordered as we'd expect:
    $\Thompson > \DynamicEpsGreedy > \DynamicGreedy$
for a sufficiently long time horizon. For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. We also plot the trajectory for instantaneous rewards (\emph{not} averaged over the previous time-periods), which provides a better view into algorithm's performance at a given time.%
\footnote{For ``instantaneous reward" at a given time $t$, we report the average (over all \MRVs) of the mean rewards at this time, instead of the average of the \emph{realized} rewards, so as to decrease the noise.}
Figure \ref{prelim_means} shows these trajectories for the Needle-in-Haystack instance; for other MAB instances the plots are similar.
We summarize this finding as follows:

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=0.35]{ec19paper/figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{ec19paper/figures/relative_nih_ts_dg}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\Thompson \geq \DynamicGreedy$. The shaded area display 95\% confidence intervals.}}
\caption{\footnotesize Relative reputation trajectory for $\Thompson$ vs $\DynamicGreedy$, on Uniform instance (left) and Needle-in-Haystack instance (right). Shaded area display 95\% confidence intervals. The relative reputation at time $t$ is the fraction of \MRVs for which, at time $t$, $\Thompson$ has a higher reputation score than $\DynamicGreedy$.}
\label{relative_rep_plots}
\end{figure}

This angle allows a more nuanced analysis of reputation costs vs. benefits under competition. Figure \ref{relative_rep_plots} (left) shows the relative reputation trajectory for $\Thompson$ vs $\DynamicGreedy$ for the Uniform instance. The relative reputation is less than $\tfrac12$ in the early rounds, meaning that $\DynamicGreedy$ has a higher reputation score in a majority of the simulations, and more than $\tfrac12$ later on. The reason is the exploration in \Thompson leads to worse decisions initially, but allows for better decisions later. The time period when relative reputation vs. \DynamicGreedy dips below $\tfrac12$ can be seen as an explanation for the competitive disadvantage of exploration. Such period also exists for the Heavy-Tail instance. However, it does not exist for the Needle-in-Haystack instance, see Figure \ref{relative_rep_plots}.%
\footnote{We see two explanations for this: $\Thompson$ identifies the best arm faster for the Needle-in-Haystack instance, and there are no ``very bad" arms to make exploration expensive in the near term.}

@@

\noindent Note that Uniform and Heavy-tail instances are exploration-disadvantaged, but Needle-in-Haystack instance is not.
